Forecasting short-term hourly Emergency Departement arrivals

### 2021-11-01 paper/paper.tex
using arrival time instead of admission in the text, put all plots in the paper

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\usepackage[top=25mm, left=30mm, right=30mm, bottom=25mm,headsep=10mm, footskip=12mm]{geometry}
\usepackage{natbib}
\bibliographystyle{plainnat}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={Forecasting hourly hospital admission},
            colorlinks=false,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls

@@

  \title{Forecasting hourly hospital admission}
    \author[University1]{Author1\corref{1}}
   \ead{email1@example.com} 
    \author[University2]{Author2\corref{2}}
   \ead{email2@example.com} 
    \author[Centre for Marketing Analytics and Forecasting, Lancaster University, UK]{Ivan Svetunkov\corref{2}}
   \ead{i.svetunkov@lancaster.ac.uk} 
      \address[University1]{Cardiff business school, 3 Colum Drive, CF10 3EU, Cardiff}
    \address[University2]{adress2}
    \address[University3]{adress3}
      \cortext[1]{Corresponding Author}
    \cortext[2]{Equal contribution}

@@

Forecasting hospital admission is a critical input to inform staffing and scheduling in hospital services to meet the needs of patients. An accurate hospital admission forecast contributes to a better decision making process regarding the resources needed to provide services to the number and type of patients requiring hospital services. This is one of the best ways to optimize staffing utilization and related costs. Most current practice to optimise personnel scheduling follows the general approach originally presented in \citet{vile2016time}, which recommends that the following steps be taken to roster employees: (i) forecast demand; (ii) convert demand forecasts into staffing requirements; (iii) schedule shifts optimally; and (iv) assign employees to shifts. An accurate demand forecasting is crucial in Accident \& Emergency and Ambulance Services in particular, to depict various courses of action that can result in massive savings in terms of patient lives \citep{rostami2020anticipating}.

Unability to match the staff with the demand results in patient overcrowding in the system which is a serious problem that causes challenging situations on patient flow \citep{derlet2002overcrowding}. Also, it is related with increasing length of stay \citep{muhammet2015forecasting}, low patient satisfaction, number of patients left the hospital without seen by staff, unexpected return visits to services, increasing health care costs, inaccuracy in electronic medical record, and reported waiting times without incurring last-minute expenses, such as overtime or supplemental staffing.

There exists a large literature on forecasting hospital admission - including emergency department and ambulance services - that use various methods to forecast annual\citep{tandberg1994time} monthly,\citep{chen2011long, mai2015predicting}. daily\citep{rostami2020anticipating, park2019144} and subdaily \citep{schweigler2009forecasting, cheng2021forecasting} admission. In this paper, we only focus on forecasting hourly hospital admission. In comparison to lower frequency time series forecasting such as monthly quarterly and yearly admission, hourly forecasts are challenging because the noise caused by random variation may overshadow any pattern in the time series.
An accurate forecast of the admission demand by hour of the day enables managers to match staff to meet anticipated patients, reconfigure units and redeploy staff. This will have many advantages for both patients, staff and the quality of provided services. Hourly forecasts are required to inform the short-term operational planning for the current and the upcoming shifts of the day. This involves the short-term decision making related to the execution of the delivery process for various health care services such as Ambulatory, Emergency, Surgical, Inpatient, Home and Residential. The combination of an hourly forecast demand, current staff being occupied (e.g incidents being attended), resource availability and waiting times at hospitals, provide information on the state of the unscheduled care system across the service. Having this full picture enables the delivery managers to focus on the areas that require intervention to enable the most effective delivery of the service to the patients. Moreover, there are often unplanned events, such as walk-in patients, extended consultation times, during a day or a week. Staff capacities may be adjusted based on the predicted demand fluctuations by using part-time, on-call nurses, staff overtime and voluntary absenteeism. That may also require to adjust the scheduling.

There are few studies that look at forecasting hourly hospital admission using historical time series data and/or predictors such as patient characteristics , climate factors , holidays. These studies use multiple approaches including Exponential Smoothing {[}REF{]}, ARIMA families {[}{]}, ARCH {[}REF{]}, Vector Autoregressive {[}REF{]}, TBATS {[}{]} and ANN{[}{]}. Hourly time series generally exhibit multiple seasonal cycles of different lengths such as hourly, daily, weekly and yearly. They may also express nonstationarity and their profile may change over time. Therefore, an appropriate forecasting model should take these features into account, to accurately predict hourly demand admissions. This is currently missing in the literature and we fill this gap by examining various forecasting models capable of considering these features. All these studies only generate the estimated future admission as a point forecast (a single number), which does not report any uncertainty around the admission. Reporting uncertainty is critical because the consequences of imperfect staffing are asymmetric. Therefore, probabilistic forecasts are necessary to make decisions that balance the cost associated with under and over staffing. This asymmetry arises because it is preferable to incur a small opportunity cost associated under utilised staff rather than lower service levels if staff levels are insufficient. The lack of probabilistic forecast is one of the main bottlenecks in the deployment of generated forecasts in the staffing and planning tools. In this paper, we produce and evaluate the whole forecast distribution of hourly hospital admission, using all forecasting models, which could be used as a risk management tool for planners and decision makers. Aditionally, datasets used in some of these studies are relatively small, short length (e.g.~time period of 1 year), which make it challenging to report the forecast accuracy using robust approaches such as time series cross validation and such results might not be reliable. Moreover, all previous publications covered in this paper fails reproducibility principles (Boylan, 2016).

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  We develop a novel methodology to forecast short-term hourly hospital admission using the family of Generalised Additive Models that accounts for i) \ldots, ii) \ldots{} iii) , \ldots{} iv)
\item
  We produce probabilistic forecast, in addition to the point estimation, that quantify uncertainties in future hospital admission and we evaluate its accuracy using \ldots;
\item
  We benchmark the accuracy of our model against appropriate models used when multiple seasonality is present, i.e.~Prophet, TBATS, Poisson Regression, and exponential smoothing state space model (ETS);
\item
  We provide data and code enabling reproduction and refinement of the proposed approach and benchmarks. The proposed approach could also be generalised to forecast hourly requirements in other services such as the number of incidents or call volums in clinical desk services.
\end{enumerate}

@@

\hypertarget{lit}{%
\section{Research background: hourly hospital admission forecasting}\label{lit}}

Other papers to review and possibly cite:
- Daily visits using Prophet \citet{park2019144}
- Daily using internet data \citet{ekstrom2015forecasting}
- Review paper: \citet{wargon2009systematic}
- Others: \citet{boyle2012predicting},
- Impact of weather \citet{parsons2011modelling}, \citet{macgregor2003effect}

hourly:
\citet{cheng2021forecasting}
\citet{rocha2021forecasting}

Table \ref{tab:summarylit} summarise studies in hourly forecasting in emergency and urgent care.

\caption{\label{tab:summarylit}Summary of studies in hourly emergency care forecasting}
\centering
\fontsize{11}{13}\selectfont
\begin{tabular}[t]{lr>{\raggedright\arraybackslash}p{6em}>{\raggedright\arraybackslash}p{4em}l>{\raggedright\arraybackslash}p{15em}>{\raggedright\arraybackslash}p{4em}ll}
\toprule
Author & Year & Variable & Horizon & Length & Method & Metric & Probabilistic & Multiple\\
\midrule
NA & NA & NA & NA & NA & NA & NA & NA & seasonality\\
McCarthy 
2008 & 2008 & ED arrivals & 24h & 2 years & Poisson log-linear regression model & 95\% CI & No & Yes\\
Asheim et al. & 2019 & ED arrivals & 3h & 5 & Poisson regression & MAPE & No & No\\
Cote et al. & 2013 & ED arrivals & 24h & 2 & Fourier regression & R\textasciicircum{}2,  Standard Error & No & No\\
Kim et al. & 2014 & Hospital admission & 4h, 24h & 3 & Linear regression; Exponential smoothing; ARIMA; GARCH; VAR & MAPE & No & No\\
Schweigler et al. & 2009 & Bed occupancy & 4h, 12h & 4 & Hourly historical average; SARIMA; Sinusoidal model with autocorrelated error & RMSE & No & No\\
Channouf et al. & 2007 & Ambulance admission & 1h, 3h, 6h, 12h, 13h, 14h, 17h, 23h, 24h & 2 & Regression & RMSE & No & No\\
Hertzum & 2017 & ED arrivals & 1,2,4,8,24 hours & 3 & linear regression; SARIMA; Naïve & MAE, MAPE, 
MASE & No & No\\
Choudhury and Urena & 2020 & ED arrivals & 1h to 24h & 4 & ARIMA; Holt-winters; TBATS; ANN & RMSE, ME & No & Yes\\
Jones et al. & 2009 & ED census & 24h & 2 & VAR;  Holt winters & MAE & No & No\\
Morzuch and Allen & 2006 & ED arrivals & 168h & 3 & Regression; ARIMA; Exponential smoothing & RMSE & No & No\\
Chase et al. & 2012 & ED CUR & 30m
1h, 2h, 4h, 8h, 12h & 1 & Binary regression & NA & No & No\\
\bottomrule
\end{tabular}
\end{table}

@@

Predicting the arrivals of accident and emergency department future patients is studied by \citet{choudhury2020forecasting} ARIMA, Holt-Winters, TBATS, and neutral network methods were implemented to forecast hourly accident and emergency department arrivals. ARIMA model was selected as the best fit model and it has provided high and acceptable hourly accident and emergency department forecasting accuracy. \citet{hertzum2017forecasting} work was mentioned in this paper. It is said that residual normality, stationarity, and autocorrelation have not been tested in \citet{hertzum2017forecasting} paper and this might be the cause of accuracy problems. However, residual normality, stationarity, and autocorrelation are tested and compared with Holt-Winters, TBATS, and neutral network methods in \citet{choudhury2020forecasting}
\citet{morzuch2006forecasting} use the Unobserved Components Model (UCM), by which each component of the time series is separately modelled as stochastic; double-seasonal exponential smoothing and standard Holt-Winters to forecast ED arrival for an horizon of 168 hours. The hourly data collected from an Emergency department in Pennsylvania showed no trend, and two seasonal cycles: a within-day and a within-week seasonal cycles. The double seasonal model recorded lower RMSEs for all the 168-hour horizons, which was expected due to the strong hourly seasonality of the time series.

\citet{mccarthy2008challenge} predict patient demand for ED services by characterising ED arrivals. Hourly arrival data of ED arrivals in the 1-year study period was deployed to forecast from 1 hour to 24 hours into the future. Authors use a Poisson log-linear regression model, including independent variables such as temporal factors (i.e., hour-of-day, day- of-week, type-of-day, season, and calendar year), patient characteristics (i.e., age, gender, insurance status, triage level, mode of arrival, and ambulance diversion status) and climatic factors (i.e., temperature and precipitation).
Then, they presente the predictiona interval accuracy of 50 and \(90\%\) intervals for the number of hourly arrivals under the Poisson assumptions. They show that the most important predictor is hour of the day and autocorrelation lag 1. However, these findings are limited to the short period of data (one year).

@@

\citet{steins2019forecasting} aimed to develop a forecasting model for predicting the number of ambulance services calls per hour and geographical areas, to support managers in decisions-making and to investigate which ones were the factors that affected the number of calls. Data collected consisted of a time and location of historical ambulance call data for three counties in Sweden and a list of explanatory factors of the area, such as socioeconomic and geographic. In order to deal with large number of zeros in the data, authors developed zero-inflated Poisson (ZIP) and zero-inflated negative binomial (ZINB) regression models. These were then compared to the currently existing forecasting system, based on moving average with seasonality weights, using ME, MAE and RMSE. Firstly, the factors affecting the number of ambulance calls were found to be the following: population in different age groups, median income, length of road, number of nightlife spots (the number of restaurants), day of the week and hour of the day. Secondly, it was found that the older population (65-100) generated more ambulance calls and that ZIP model performs better than the current model. However, the improvement provided by the more advance model was not much greater than the one provided by the existing model. Authors suggested that it could be because the population, which was used as an independent variable in both models, was so dominant compared to the other factors.
Moreover, both models either underestimated or overestimated the number of calls. Authors suggested that the inability to capture a positive trend resulted in underestimation, and the opposite was due to a negative trend. Therefore, authors recommended that further research should add certain temporal variables able to capture the trend.

According to the studies mentioned earlier, it can be said that the existing studies have shown complications in forecasting hourly patient accident and emergency department visits and the application of forecasting hourly patients visits is not well established. Some of the studies said that the accuracy of hourly accident and emergency department forecasting model is low compared to other longer forecasting intervals like daily forecast \citep{boyle2012predicting, hertzum2017forecasting}. However, some studies mentioned that the accuracy of accident and emergency department hourly forecast is at the acceptable level \citep{choudhury2020forecasting, mccarthy2008challenge, schweigler2009forecasting}.

There are few limitations in the literature which encourage us to undertake this research and propose examine different forecasting approaches to deal with them. These limitations are summarised as follows : (i) Current approaches to forecast hourly hospital admission do not fully consider the feature of data such as multiple seasonal cycles and changing profile over time; (ii) Almost all research studies produce point forecast and consequently report only point forecast accuracy. There is a lack of studies presenting probabilistic forecast of hourly hospital admission that better represent uncertainty of future admissions, providing an holistic picture of future admission for a planner; (iii) most studies are not reproducible as it is almost impossible to reproduce forecasting models and results; (iv) studies are limited in terms of the length of historical data used for training purposes and forecast performance evaluation and (v) some studies in this area lack a rigorous experimental design, i.e.~there is no benchmark method nor is forecast accuracy reported.

@@

Figure \ref{fig:hourly-plot-ridge}

@@

Figure \ref{fig:hourly-plot}

@@

{\centering \includegraphics[width=0.8\linewidth]{paper_files/figure-latex/hourly-plot-1}

@@

Let \(y_t\) be the number of attendances in time period \(t\) and indicate with \(F_t(y_t)\) its predictive cumulative probability distribution. In a distributional regression context, \(F_t(y_t)\) is modelled via a parametric model, \(F(y_t|\bm \theta_t)\), where \(\bm \theta_t\) is an \(m\)-dimensional vector of parameters. In a GAMLSS framework \citet{Rigby2005} the elements \(j=1,...,m\) of \(\bm \theta_t\) are modelled via

@@

Probabilistic forecasts are evaluated following the principle of \emph{sharpness subject to calibration}, meaning that the sharper forecast is prefered provided that it is calibrated. Mis-calibrated forecasts are unsuitable for use in decision-making so should be excluded. Calibration is evaluated visually in \ref{fig-quantilebias}, which highlights a systematic negative bias accross all probability levels in many models, with only the truncated normal and \(t\) family GAMLSS models (NOtr-1, NOtr-2, Ttr-2) and iETS models showing good calibration across most probability levels. Notably, both benchmarks exhibit negative quantile bias as they struggle to capture the long term trend of increasing attendance.

@@

{\centering \includegraphics[width=0.9\linewidth]{../results/QuantileBias}

@@

\caption{Quantile ...}\label{fig:fig-quantilebias}
\end{figure}

@@

\caption{\label{tab:tab-results-qbias}Forecast evluation for all methods on test data in decreasing order by Quantile Bias}
\centering
\fontsize{11}{13}\selectfont
\begin{tabular}[t]{r}
\toprule
Quantile Bias\\
\midrule
0.1047874\\
0.0557392\\
0.0194389\\
0.0600650\\
0.0540740\\
0.0375705\\
0.0098936\\
0.0118522\\
0.0372137\\
0.0373884\\
0.0523937\\
0.0929416\\
0.0140252\\
0.1862190\\
0.0104673\\
0.0104673\\
0.0643451\\
0.0855702\\
\bottomrule
\end{tabular}
\end{table}

@@

{\centering \includegraphics[width=0.9\linewidth]{../results/Skill_rel2bench_reduced}

@@

\caption{Skill score ...}\label{fig:fig-skill}
\end{figure}

>>>>

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\usepackage[top=25mm, left=30mm, right=30mm, bottom=25mm,headsep=10mm, footskip=12mm]{geometry}
\usepackage{natbib}
\bibliographystyle{plainnat}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={Forecasting short-term hourly Emergency Departement arrivals},
            colorlinks=false,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls

@@

  \title{Forecasting short-term hourly Emergency Departement arrivals}
    \author[University1]{Author1\corref{1}}
   \ead{email1@example.com} 
    \author[School of Mathematics and Statistics, University of Glasgow, UK]{Jethro Browell\corref{2}}
   \ead{jethro.browell@glasgow.ac.uk} 
    \author[Centre for Marketing Analytics and Forecasting, Lancaster University, UK]{Ivan Svetunkov\corref{2}}
   \ead{i.svetunkov@lancaster.ac.uk} 
      \address[University1]{Cardiff business school, 3 Colum Drive, CF10 3EU, Cardiff}
    \address[University2]{adress2}
    \address[University3]{adress3}
      \cortext[1]{Corresponding Author}
    \cortext[2]{Equal contribution}

@@

Forecasting Emergency Department (ED) arrivals is a critical input to inform staffing and scheduling decisions to meet the needs of patients. An accurate ED demand forecast contributes to a better decision making process regarding the resources needed to provide services to the number and type of patients requiring hospital services. This is one of the best ways to optimize staffing utilization and related costs. Most current practice to optimise personnel scheduling follows the general approach originally presented in \citet{vile2016time}, which recommends that the following steps be taken to roster employees: (i) forecast demand; (ii) convert demand forecasts into staffing requirements; (iii) schedule shifts optimally; and (iv) assign employees to shifts. An accurate demand forecasting is crucial in ED services to depict various courses of action that can result in massive savings in terms of patient lives. Inability to match the staff with the demand results in patient overcrowding in the system which is a serious problem that causes challenging situations on patient flow \citep{derlet2002overcrowding}. Also, it is related with increasing length of stay \citep{muhammet2015forecasting}, low patient satisfaction, number of patients left the hospital without seen by staff, unexpected return visits to services, increasing health care costs, inaccuracy in electronic medical record, and reported waiting times without incurring last-minute expenses, such as overtime or supplemental staffing \citep{rostami2020anticipating}.

There exists a large literature on forecasting ED arrivals that use various methods to forecast annual\citep{tandberg1994time} monthly,\citep{chen2011long, mai2015predicting}. daily\citep{rostami2020anticipating, park2019144} and subdaily \citep{schweigler2009forecasting, cheng2021forecasting} arrivals. In this paper, we only focus on forecasting hourly arrivals. In comparison to lower frequency time series forecasting such as monthly, quarterly and yearly, hourly forecasts are challenging because the noise caused by random variation may overshadow any pattern in the time series.

An accurate demand forecast by hour of the day enables planners to match staff to meet anticipated patients, reconfigure units and redeploy staff. This will have many advantages for both patients, staff and the quality of provided services. Hourly forecasts are required to inform the short-term operational planning for the current and the upcoming shifts of the day. This involves the short-term decision making related to the execution of the delivery process in ED.
The combination of an hourly forecast demand, current staff being occupied, resource availability and waiting times at ED, provide information on the state of the unscheduled care system across the service. Having this full picture enables the delivery managers to focus on the areas that require intervention to enable the most effective delivery of the service to the patients. Moreover, there are often unplanned events, such as walk-in patients, extended consultation times, during a day or a week. Staff capacities may be adjusted based on the predicted demand fluctuations by using part-time, on-call nurses, staff overtime and voluntary absenteeism. That may also require to adjust the scheduling.

There are few studies that look at forecasting hourly arrivals in ED and other hospital services using historical time series data and/or predictors such as patient characteristics , climate factors , holidays. These studies use multiple approaches including Exponential Smoothing {[}REF{]}, ARIMA families {[}{]}, ARCH {[}REF{]}, Vector Autoregressive {[}REF{]}, TBATS {[}{]} and ANN{[}{]}. Hourly time series generally exhibit multiple seasonal cycles of different lengths such as hourly, daily, weekly and yearly. They may also express nonstationarity and their profile may change over time. Therefore, an appropriate forecasting model should take these features into account, to accurately predict hourly demand admissions. This is currently missing in the literature and we fill this gap by examining various forecasting models capable of considering these features. All these studies only generate the estimated future arrivals as a point forecast (a single number), which does not include any uncertainty around the number of arrivals. However, reporting uncertainty is critical because the consequences of imperfect staffing are asymmetric. Therefore, probabilistic forecasts are necessary to make decisions that balance the cost associated with under and over staffing. This asymmetry arises because it is preferable to incur a small opportunity cost associated under utilised staff rather than lower service levels if staff levels are insufficient. The lack of probabilistic forecast is one of the main bottlenecks in the deployment of generated forecasts in the staffing and planning tools. In this paper, we produce and evaluate the whole forecast distribution of hourly ED arrivals, using all forecasting models, which could be used as a risk management tool for planners and decision makers. Moreover, datasets used in some of these studies are relatively small, short length (e.g.~time period of 1 year), which make it challenging to report the forecast accuracy using robust approaches such as time series cross validation and such results might not be reliable. Moreover, all previous publications covered in this paper fails reproducibility principles.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  We develop a novel methodology to forecast short-term hourly hospital admission using the family of Generalised Additive Models that accounts for i) \ldots, ii) \ldots{} iii) , \ldots{} iv) (Jethro to complete)
\item
  We produce probabilistic forecast, in addition to the point estimation, that quantify uncertainties in future hospital admission and we evaluate its accuracy using \ldots;
\item
  We benchmark the accuracy of our model against appropriate models used when multiple seasonality is present, i.e.~Prophet, TBATS, Poisson Regression, and exponential smoothing state space model (ETS);
\item
  We provide data and code enabling reproduction and refinement of the proposed approach and benchmarks. The proposed approach could also be generalized to forecast hourly requirements in other services such as the number of incidents or call volumes in clinical desk services.
\end{enumerate}

@@

\hypertarget{lit}{%
\section{Research background: hourly ED forecasting}\label{lit}}

There are many studies that employs models to forecast the Emergency Department needs such as admission and arrivals. The time granularity considered by these studies spans from hourly to yearly. However, given the focus of the paper, we only focus on studies on hourly ED forecasting. Table \ref{tab:summarylit} summarize studies in hourly forecasting in the emergency department.

\begin{landscape}\begin{table}[!h]

\caption{\label{tab:summarylit}Summary of studies in hourly Emergency Department forecasting}
\centering
\resizebox{\linewidth}{!}{
\fontsize{11}{13}\selectfont
\begin{tabular}[t]{lr>{\raggedright\arraybackslash}p{6em}>{\raggedright\arraybackslash}p{4em}l>{\raggedright\arraybackslash}p{15em}>{\raggedright\arraybackslash}p{4em}ll}
\toprule
Author & Year & Variable & Horizon & Length & Method & Metric & Probabilistic & Multiple seasonality\\
\midrule
McCarthy 
2008 & 2008 & ED arrivals & 24h & 2 years & Poisson log-linear regression model & 95\% CI & No & Yes\\
Asheim et al. & 2019 & ED arrivals & 3h & 5 & Poisson regression & MAPE & No & No\\
Cote et al. & 2013 & ED arrivals & 24h & 2 & Fourier regression & R\textasciicircum{}2,  Standard Error & No & No\\
Kim et al. & 2014 & Hospital admission & 4h, 24h & 3 & Linear regression; Exponential smoothing; ARIMA; GARCH; VAR & MAPE & No & No\\
Schweigler et al. & 2009 & Bed occupancy & 4h, 12h & 4 & Hourly historical average; SARIMA; Sinusoidal model with autocorrelated error & RMSE & No & No\\
Channouf et al. & 2007 & Ambulance admission & 1h, 3h, 6h, 12h, 13h, 14h, 17h, 23h, 24h & 2 & Regression & RMSE & No & No\\
Hertzum & 2017 & ED arrivals & 1,2,4,8,24 hours & 3 & linear regression; SARIMA; Naïve & MAE, MAPE, 
MASE & No & No\\
Choudhury and Urena & 2020 & ED arrivals & 1h to 24h & 4 & ARIMA; Holt-winters; TBATS; ANN & RMSE, ME & No & Yes\\
Jones et al. & 2009 & ED census & 24h & 2 & VAR;  Holt winters & MAE & No & No\\
Morzuch and Allen & 2006 & ED arrivals & 168h & 3 & Regression; ARIMA; Exponential smoothing & RMSE & No & No\\
Chase et al. & 2012 & ED CUR & 30m
1h, 2h, 4h, 8h, 12h & 1 & Binary regression & NA & No & No\\
\bottomrule
\end{tabular}}
\end{table}
\end{landscape}

@@

Predicting the arrivals of an ED future arrivals is also studied by \citet{choudhury2020forecasting}. ARIMA, Holt-Winters, TBATS, and neutral network methods were implemented to forecast hourly accident and emergency department arrivals. ARIMA model was selected as the best fit model. Authors claimed that ARIMA has provided high and acceptable hourly ED forecasting accuracy. \citet{hertzum2017forecasting} work was mentioned in this paper. This result is surprising given the existence of multiple seasonality in the hourly dataset.

\citet{morzuch2006forecasting} used the Unobserved Components Model (UCM), by which each component of the time series is separately modeled as stochastic. Double-seasonal exponential smoothing and standard Holt-Winters are used to forecast ED arrival for an horizon of 168 hours. The hourly data collected from an ED in Pennsylvania showed no trend, and two seasonal cycles: a within-day and a within-week seasonal cycles. The double seasonal model recorded lower RMSEs for all the 168-hour horizons, which was expected due to the strong hourly seasonality of the time series.

\citet{mccarthy2008challenge} employed a Poisson log-linear regression model, including independent variables such as temporal factors (i.e., hour-of-day, day- of-week, type-of-day, season, and calendar year), patient characteristics (i.e., age, gender, insurance status, triage level, mode of arrival, and ambulance diversion status) and climatic factors (i.e., temperature and precipitation) to forecast patient demand for ED services. Hourly arrival data of ED arrivals in the 1-year study period was deployed to forecast from 1 hour to 24 hours into the future. Authors . They present the prediction interval accuracy of \(50\%\) and \(90\%\) intervals for the number of hourly arrivals under the Poisson assumptions. They show that the most important predictor is hour of the day and autocorrelation lag 1. However, these findings are limited to the short number of observations (only one year of historical data).

@@

According to the studies mentioned earlier, it can be said that the existing studies have shown complications in forecasting hourly patient accident and emergency department visits and the application of forecasting hourly patients visits is not well established. Some of the studies said that the accuracy of hourly accident and emergency department forecasting model is low compared to other longer forecasting intervals like daily forecast \citep{boyle2012predicting, hertzum2017forecasting}. However, some studies mentioned that the accuracy of ED hourly forecast is at the acceptable level \citep{choudhury2020forecasting, mccarthy2008challenge, schweigler2009forecasting}.

There are few limitations in the literature which encourage us to undertake this research and propose examine different forecasting approaches to deal with them. These limitations are summarised as follows : (i) Current approaches to forecast hourly ED arrivals do not fully consider the feature of data such as multiple seasonal cycles and changing profile over time; (ii) Almost all research studies produce point forecast and consequently report only point forecast accuracy. There is a lack of studies presenting probabilistic forecast of hourly ED arrivals that better represent uncertainty of future admissions, providing an holistic picture of future demand for a planner; (iii) most studies are not reproducible, as it is almost impossible to reproduce forecasting models and results from the studies; (iv) studies are limited in terms of the length of historical data used for training purposes and forecast performance evaluation and (v) some studies in this area lack a rigorous experimental design, i.e.~there is no benchmark method nor is forecast accuracy reported.

@@

Data used in the study comprised counts of patients' arrival times at one of the largest ED units in the UK between April 2014 and February 2019, extracted from the ED administrative database of the hospital. We aggregated the patients' arrival times to obtain hourly arrivals, which is used for empirical evaluation in this study.

Figure \ref{fig:hourly-plot-ridge} illustrates the distribution of arrivals for each hour of the day and the day of the week. Although the data is noisy but it reveals some systematic structures.

It is clear that the number of arrivals has a sub-daily structure. The ED arrivals decreases between mid-night and early morning and then increases until the evening and then decreases again. It is also clear that ED service gets systematically more visits on Monday between 8 a.m. and 5 p.m. Moreover, The number of arrivals around mid-night is slightly higher for Saturday and Sunday.

@@

Figure \ref{fig:hourly-plot} also highlights that there are many ouliers for almost every hour of the day that may affec the accuracy of forecasting methods, some of these outliers are might be related to holidays and special events. It is also clear that arrivals are less volatile between mid-night and early morning.

We propose a forecasting model that accounts for the systematic structure of the data.

@@

{\centering \includegraphics[width=0.7\linewidth]{paper_files/figure-latex/hourly-plot-1}

@@

If we assume that our predictive distribution follows a given parametric distribution, the forecasting task becomes ones of predicting the future values of that distribution's parameters. Generalised Additive Models for Location, Scale and Shape (GAMLSS) are distributional regression models where the parameters of the assumed distribution may be modelled as additive functions of explanatory variables. This provides a powerful and flexible framework for probabilistic forecasting, provided that a suitable distribution and additive structures can be found. In practice, this means experimenting with various distributions and evaluating their suitability using available training data.
Let \(y_t\) be the number of attendances in time period \(t\) and indicate with \(F_t(y_t)\) its predictive cumulative probability distribution. In a distributional regression context, \(F_t(y_t)\) is modelled via a parametric model, \(F(y_t|\bm \theta_t)\), where \(\bm \theta_t\) is an \(m\)-dimensional vector of parameters. In a GAMLSS framework \citet{Rigby2005} the elements \(j=1,...,m\) of \(\bm \theta_t\) are modelled via

@@

Probabilistic forecasts are evaluated following the principle of \emph{sharpness subject to calibration}, meaning that the sharper forecast is prefered provided that it is calibrated. Mis-calibrated forecasts are unsuitable for use in decision-making so should be excluded. Calibration is evaluated visually in \ref{quantile-bias}, which highlights a systematic negative bias accross all probability levels in many models, with only the truncated normal and \(t\) family GAMLSS models (NOtr-1, NOtr-2, Ttr-2) and iETS models showing good calibration across most probability levels. Notably, both benchmarks exhibit negative quantile bias as they struggle to capture the long term trend of increasing attendance.

@@

{\centering \includegraphics[width=0.9\linewidth]{paper_files/figure-latex/Pinball-1}

@@

\caption{pinball ...}\label{fig:Pinball}
\end{figure}

{\centering \includegraphics[width=0.9\linewidth]{paper_files/figure-latex/lead-time-rel-1}

\caption{pinball ...}\label{fig:lead-time-rel}
\end{figure}

{\centering \includegraphics[width=0.9\linewidth]{paper_files/figure-latex/Reliability-1}

\caption{Reliability ...}\label{fig:Reliability}
\end{figure}

{\centering \includegraphics[width=0.9\linewidth]{paper_files/figure-latex/quantile-bias-1}

\caption{Quantile ...}\label{fig:quantile-bias}
\end{figure}

@@

@@

{\centering \includegraphics[width=0.7\linewidth]{paper_files/figure-latex/Skill-rel2bench-reduced-1}

@@

\caption{Skill score ...}\label{fig:Skill-rel2bench-reduced}
\end{figure}

{\centering \includegraphics[width=0.7\linewidth]{paper_files/figure-latex/rmse-1}

\caption{RMSE ...}\label{fig:rmse}
\end{figure}

{\centering \includegraphics[width=0.7\linewidth]{paper_files/figure-latex/lead-time-rmse-1}

\caption{RMSE ...}\label{fig:lead-time-rmse}
\end{figure}

{\centering \includegraphics[width=0.7\linewidth]{paper_files/figure-latex/time-1}

\caption{Running time ...}\label{fig:time}
\end{figure}
